\section{Constant and variables}

Within a program, declared symbols are identified by a name. Names (also called
identifiers) follow specific grammar rules and are case-sensitive (i.e., upper
and lower case are important). Identifiers can contain numbers, letters, and
underscores, but they must contain at least one letter, and numbers are always
preceded by at least one letter. An identifier can starts with an arbitry number
of underscores. For example, the following identifier is valid \token{\_\_id1\_\_},
but the identifier \token{\_\_1id\_\_} is not.

\mynotebox{
  The formal grammar of a identifier is the following: \token{(\_)* [a-Z]([a-Z]|[0-9]|\_)*}
}

Some names that respect the grammar of identifier cannot be used as identifier,
as they are reserved by the language to define specific statements. These
identifiers are called keywords, and the complete list of reserved keywords is
given below. The meaning of these keywords will be explained through the course
of this book. Keyword list: \token{alias}, \token{assert}, \token{atomic},
\token{break}, \token{cast}, \token{catch}, \token{class}, \token{const},
\token{copy}, \token{cte}, \token{dcopy}, \token{def}, \token{dg}, \token{dmut},
\token{do}, \token{else}, \token{enum}, \token{expand}, \token{extern},
\token{false}, \token{for}, \token{fn}, \token{future}, \token{if},
\token{impl}, \token{in}, \token{is}, \token{lazy}, \token{let}, \token{loop},
\token{macro}, \token{match}, \token{mod}, \token{move}, \token{mut},
\token{none}, \token{null}, \token{of}, \token{panic}, \token{\_\_pragma},
\token{record}, \token{union}, \token{ref}, \token{return}, \token{spawn},
\token{static}, \token{entity}, \token{throw}, \token{throws}, \token{trait},
\token{true}, \token{typeof}, \token{\_\_test}, \token{unsafe}, \token{use},
\token{\_\_version}, \token{while}, \token{with}.


